Towards a wide-coverage grammar for Swedish using GF

### 2011-12-28 doc/report/report.tex
report, modification to passive
report: abstract, acknowledgments

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Towards a wide-coverage grammar for Swedish} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

\abstract{
%This is the report of the work of expanding and enhancing a GF grammar and preparing
%it for parsing free Swedish. The grammar is a result of combining resources,
%we have extended the current GF grammar to adapt it for Swedish specifically so
%that it can parse an language specific constructions. 
%We show how dependent types can be used to give a neater description of
%syntax.
%We have further completed a method for importing the electronic 
%lexicon Saldo to a GF format and established a connection to the treebank Talbanken.
%The imported lexicon consists of more than 100 000 words.

@@

\newpage

\section*{Acknowledgments}
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage

\tableofcontents
\newpage
\chapter{Introduction}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism,
developed for multilinguality. It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has so far been successfully used for 
controlled languages \cite{cnl}, but recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
This project focuses on implementing a wide-coverage Swedish grammar, which also
describes a parser accepting all sentences that can be expressed by the grammar.

@@

During this project, the s-passive was added as the standard case.
Periphrastic passive is still allowed, but as an alternative rather
than the default.
The grammar further allows not only V2, but all verb phrases that misses an object, to 
form passives:
\begin{verbatim}
PassVPSlash : VPSlash -> VP ;
              ta      -> togs
              erbjöd  -> erbjöds
\end{verbatim}
\enumsentence{\textbf{First place in two-place verbs}\\\shortex{3}
{Jobbet & erbjöds & henne }
{\emph{the job} & \emph{offered+\textbf{s}}&\emph{ henne}}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{\textbf{Second place in two-place verbs}\\\shortex{3}
{Hon & erbjöds & jobbet}
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\textbf{V2A verb}\\\shortex{3}
{Huset & målades & rött}
{\emph{the house} & \emph{painted+\textbf{s}}& \emph{red}}
{``The house was painted red"}}
\label{ex:passV2A}

>>>>

\begin{document}
% ska gå att förstå och locka till läsning
%ska tydligt beskriva vad rapporten handlar om
%ska innehålla vettiga sökord
\title{Towards a wide-coverage grammar for Swedish using GF} %A nice titleScaling up a GF grammar and preparing it for cool stuff for Swedish}
\author{Malin Ahlberg}
\maketitle
\newpage

\abstract{
This thesis describes the work to achieve a covering Swedish grammar implemented
in GF -- to model natural language by functional programming. 
The idea has been to combine existing language technology resources with
new techniques, and the aim to obtain a system for analysing written text.
To reach this goal, problems of computational as well as linguistic nature had to be solved.
\\The thesis explains the methods and technologies used and further inspects the possibilities
of reaching our long-term goal; to accomplish a parser for unrestricted Swedish, by
combining our rule-based grammar with statistical methods.
\\Our contribution is a wide-coverage GF lexicon, a translation of a Swedish treebank
to GF notation and we also give an implementation Swedish grammar. The grammar is based
on the multilingual abstract syntax given in GF resource library, and now
extends it to cover constructions specific to Swedish.
We further give an example of the advantage of using dependent types when
describing grammar and syntax, in this case reflexive pronouns.

%This is the report of the work of expanding and enhancing a GF grammar and preparing
%it for parsing free Swedish. The grammar is a result of combining resources,
%we have extended the current GF grammar to adapt it for Swedish specifically so
%that it can parse an language specific constructions. 
%We show how dependent types can be used to give a neater description of
%syntax.
%We have further completed a method for importing the electronic 
%lexicon Saldo to a GF format and established a connection to the treebank Talbanken.
%The imported lexicon consists of more than 100 000 words.

@@

\newpage
\pagestyle{empty}
\section*{Acknowledgments}
\vspace{5mm}
Many people have helped me during this work and made this thesis possible.
I would first of all like to thank
Center of Language Technology, that has funded the project.\\

Further, thanks to my excellent 
%(~~devoted~~) 
supervisor Ramona Enache for all her help and guidence 
%(encouragement)
in every phase and all aspects of the work.
Thanks to Elisabet Engdahl for sharing her seemingly unlimited knowledge
of Swedish grammar. She has also has acted as a second supervisor, and given me
very helpful comments and suggestions. Thanks to Aarne Ranta for all his great
ideas and for letting me do this project.

I am also grateful to Krasimir Angelov, Markus Forsberg, Peter Ljunglöf, Lars Borin
and many others who have contributed with ideas and inspiration, and for their
interest in this work.\\

Finally, I would also like to thank my friends and family. Special thanks to Dan Rosén
for all his support, advice and patience and, most importantly, for being such a good friend. %for having listened to 
%my monologues about the daily progress. 
% Ramona, Aarne, Elisabet, Markus, Peter?, Lars, Dan, Krasimir etc
% Funding!

% check references! lncs is good maybe. lnai
\newpage
%\pagestyle{empty}
%\cleardoublepage
\tableofcontents
%
\addtocontents{toc}{\protect\thispagestyle{empty}}
%\pagestyle{empty}
%\cleardoublepage
\newpage
\setcounter{page}{1} %reset the page counter
\chapter{Introduction}
% more intro to gf. dependently typed...high-level formalism
Grammatical Framework \cite{gfbok} is a dependently typed grammar formalism,
developed for multilinguality. It is based on Martin-Löf type theory which allows
reasoning within the programming language.
GF has so far been successfully used for 
controlled languages \cite{cnl}, but recent experiments have showed
that it is also possible to use the framework for parsing free language \cite{patent}.
% >> although the success rate is going steeply down when using uncontrolled nl. Better with annotated.
% --> sentence about connection GF - grammar based parser.
This project focuses on implementing a wide-coverage Swedish grammar, which also
describes a parser accepting all sentences that can be expressed by the grammar.

@@

During this project, the s-passive was added as the standard case.
Periphrastic passive is still allowed, but as an alternative rather
than the default.
The grammar further allows not only V2, but all verb phrases that misses an object, to 
form passives:
\begin{verbatim}
PassVP : VPSlash -> VP ;
         ta      -> togs
         erbjöd  -> erbjöds
\end{verbatim}
\enumsentence{\textbf{First place in two-place verbs}\\\shortex{3}
{Jobbet & erbjöds & henne }
{\emph{the job} & \emph{offered+\textbf{s}}&\emph{ henne}}
{``The job was offered to her"}}
\label{ex:passV32}
\enumsentence{\textbf{Second place in two-place verbs}\\\shortex{3}
{Hon & erbjöds & jobbet}
{\emph{she} &\emph{ offered+\textbf{s}}&\emph{ the job}}
{``She was offered the job"}}
\label{ex:passV33}
\enumsentence{\textbf{V2A verb}\\\shortex{3}
{Huset & målades & rött}
{\emph{the house} & \emph{painted+\textbf{s}}& \emph{red}}
{``The house was painted red"}}
\label{ex:passV2A}
